---
title: "CLL WGS analysis"
author: "Stacy Hung"
date: "2021/02/08"
output: html_document
---

This script concerns the initial analysis of "quads" of patient samples where each quad is comprised of:
1. Sorted tumor sample
2. Sorted tumor sample (biological replicate)
3. Buccal sample (normal, with potential contamination)
4. Sorted normal sample

For this pilot analysis, we have the completed quad for:
1. MB0141CLL
2. MB0291CLL

## Load in libraries

```{r}
library(ggplot2)
library(tidyverse)
library(DataCombine)  # find and replace
library(reshape2)     # colsplit
library(RColorBrewer)
library(gridExtra)
library(cowplot)
```

## Read in metrics

See https://broadinstitute.github.io/picard/picard-metric-definitions.html#CollectWgsMetrics.WgsMetrics for details

```{r}
metrics <- read.table("~/Documents/CLL_WGS/metrics/all.wgs_metrics.txt", sep = "\t", header = TRUE)

# extract patient ID
metrics <- separate(
  metrics,
  sampleID,
  into = c("patient", "tissue_type"),
  sep = "-",
  remove = FALSE
)

# get mean coverage
metrics.avg_coverage <- mean(metrics$MEAN_COVERAGE)
# convert from wide to long
metrics.long <- gather(metrics, metric, value, "MEAN_COVERAGE":"SD_COVERAGE")

# sort by decreasing order of metric
data.sorted <- metrics.long %>%
  dplyr::arrange(metric, value) %>%
  dplyr::select(sampleID, patient, tissue_type, metric, value)

# Plot MEAN_COVERAGE and SD_COVERAGE
p.sorted <- ggplot(data.sorted, aes(x = sampleID, y = value, fill=tissue_type)) + 
      geom_bar(stat = "identity", width = 0.8) + 
      facet_wrap( ~ metric, scales = "free") + 
      ylab("") + xlab("") +
      theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 10))

pdf("~/Documents/CLL_WGS/figures/barplot-mean_and_SD_coverage.pdf", width = 10, height = 5)
p.sorted
dev.off()

mean(metrics$PCT_EXC_MAPQ) # fraction of bases removed due to low mapping quality
mean(metrics$PCT_EXC_BASEQ) # fraction of bases removed due to low base quality
mean(metrics$PCT_EXC_DUPE) # fraction of bases removed due to duplicates
mean(metrics$PCT_EXC_TOTAL) # total bases removed based on all filters
```

Samples: `r unique(metrics$sampleID)`

Average coverage (n=`r length(unique(metrics$sampleID))`): `r round(metrics.avg_coverage)`X


## Define samples to be excluded and global variables

```{r}
# mappings for 3-letter amino acids to 1-letter amino acid codes	
AA.replaces <- data.frame(from = c("Ala", "Arg", "Asn", "Asp", "Asx", "Cys", "Glu", "Gln", "Glx", "Gly", "His", 	
                                   "Ile", "Leu", "Lys", "Met", "Phe", "Pro", "Ser", "Thr", "Trp", "Tyr", "Val"), 	
                          to = c("A", "R", "N", "D", "B", "C", "E", "Q", "Z", "G", "H", 	
                                 "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V"))	

# filter for protein coding biotypes
# See https://github.com/pcingola/SnpEff/blob/master/src/main/java/org/snpeff/interval/BioType.java
biotype.keeps <- c("IG_C_gene", "IG_D_gene", "IG_gene", "IG_J_gene", "IGL_V_gene", "IG_M_gene", "IG_V_gene",
                   "IG_Z_gene", "nonsense_mediated_decay", "nontranslating_CDS", "non_stop_decay",
                   "polymorphic_pseudogene", "TR_C_gene", "TR_D_gene", "TR_J_gene", "protein_coding")

# define thresholds for all samples EXCEPT for tumors paired to LMD normals
MAX_SOMATIC_PVAL = 0.05
MIN_VAR_READS_TUMOR = 3
MIN_VAF = 5
```

Strategy for assessing ability of buccals to serve as normals:

1. Compare variant overlap between T vs. buccal and T vs. sorted normal
2. Assess "purity" of buccal samples by looking at buccal vs. sorted normal
3. Do we see high numbers of TP53 mutations?


## ####### ##
## VARSCAN ##
## ####### ##

```{r}
# T vs. buccal
snvs.T_vs_buccal <- read.table("~/Documents/CLL_WGS/varScan/T_vs_buccalN/all_snvs.snpEff_canonical.dbsnp_annotated.cosmic_annotated.effect_filtered.biotype_filtered.txt", sep = "\t", header = TRUE, fill = TRUE)
indels.T_vs_buccal <- read.table("~/Documents/CLL_WGS/varScan/T_vs_buccalN/all_indels.snpEff_canonical.dbsnp_annotated.cosmic_annotated.effect_filtered.biotype_filtered.txt", sep = "\t", header = TRUE, fill = TRUE)

# T / buccal vs. sorted
snvs.T_vs_sorted <- read.table("~/Documents/CLL_WGS/varScan/T_vs_sortedN/all_snvs.snpEff_canonical.dbsnp_annotated.cosmic_annotated.effect_filtered.biotype_filtered.txt", sep = "\t", header = TRUE, fill = TRUE)
indels.T_vs_sorted <- read.table("~/Documents/CLL_WGS/varScan/T_vs_sortedN/all_indels.snpEff_canonical.dbsnp_annotated.cosmic_annotated.effect_filtered.biotype_filtered.txt", sep = "\t", header = TRUE, fill = TRUE)

snvs <- rbind(snvs.T_vs_buccal, snvs.T_vs_sorted)
rm(snvs.T_vs_buccal, snvs.T_vs_sorted)
indels <- rbind(indels.T_vs_buccal, indels.T_vs_sorted)
rm(indels.T_vs_buccal, indels.T_vs_sorted)
```

## VarScan - Processing data for filtering

```{r}
# specify type of mutation
indels$type <- "indel"
snvs$type <- "snv"

# combine snvs and indels
calls <- rbind(indels, snvs)
rm(indels)
rm(snvs)

# rename columns	
colnames(calls) <- c("sample", "chr", "pos", "external_id", "ref", "alt", "filter", 
                             "gt_normal", "gq_normal", "normal.depth", "normal.ref_reads", "normal.var_reads",
                             "normal.allele_freq", "normal.depth4", "tumor.gt", "tumor.gq", "tumor.depth",
                             "tumor.ref_reads", "tumor.var_reads", "tumor.allele_freq", "tumor.depth4", 
                             "depth", "somatic", "somatic_status", "gpv", "somatic_p_value", "allele", 
                             "effect", "impact", "gene", "feature", "feature_id", "biotype", "exon_rank", 
                             "hgvs_cdna", "hgvs_protein", "cds_pos", "cds_len", "AA_pos", "AA_len", "type")

# create a tumor_id / normal id column based on the sample id (format is <tumor_id>_<normal_id>)
calls$tumor.id <- gsub("(.*)\\_(.*)","\\1", calls$sample)
calls$normal.id <- gsub("(.*)\\_(.*)","\\2", calls$sample)

# remove "%" from the allele frequency	
calls$normal.allele_freq <- gsub("(.*)\\%(.*)","\\1", calls$normal.allele_freq)	
calls$tumor.allele_freq <- gsub("(.*)\\%(.*)","\\1", calls$tumor.allele_freq)	

# use 1-letter amino acid codes
calls <- FindReplace(data = calls, 	
                                    Var = "hgvs_protein",	
                                    replaceData = AA.replaces,	
                                    from = "from",	
                                    to = "to",	
                                    exact = FALSE)	

```

#table(calls$sample)

MB0141CLL-tumor1_MB0141CLL-buccal MB0141CLL-tumor2_MB0141CLL-buccal 
                            62710                             62005 
MB0291CLL-tumor1_MB0291CLL-buccal MB0291CLL-tumor2_MB0291CLL-buccal 
                            32295                             32173 
MB0141CLL-buccal_MB0141CLL-normal MB0141CLL-tumor1_MB0141CLL-normal 
                            56523                             63408 
MB0141CLL-tumor2_MB0141CLL-normal MB0291CLL-buccal_MB0291CLL-normal 
                            62616                             28531 
MB0291CLL-tumor1_MB0291CLL-normal MB0291CLL-tumor2_MB0291CLL-normal 
                            62298                             62170 

## VarScan - Filtering

```{r}
# VarScan-specific filters - somatic p-value + strand bias
# --------------------------------------------------------

# filter for somatic variants
calls <- filter(calls, calls$somatic == "true") # keep 4196 of 77856 variants

# filter for somatic p-value depending on the LMD or not
calls <- filter(calls,as.numeric(somatic_p_value) < MAX_SOMATIC_PVAL) # keep 4061 of 4196 variants
calls$somatic_p_value <- NULL # no longer needed

#>table(calls$sample)
#MB0141CLL-tumor1_MB0141CLL-buccal MB0141CLL-tumor2_MB0141CLL-buccal 
#                              451                               426 
#MB0291CLL-tumor1_MB0291CLL-buccal MB0291CLL-tumor2_MB0291CLL-buccal 
#                              248                               256 
#MB0141CLL-buccal_MB0141CLL-normal MB0141CLL-tumor1_MB0141CLL-normal 
#                             1083                               219 
#MB0141CLL-tumor2_MB0141CLL-normal MB0291CLL-buccal_MB0291CLL-normal 
#                              220                               544 
#MB0291CLL-tumor1_MB0291CLL-normal MB0291CLL-tumor2_MB0291CLL-normal 
#                              211                               258 

# extract depth4 fields in order to apply strand filter
calls <-	
  separate(data = calls,	
           col = tumor.depth4,	
           into = c("tumor.ref_reads.forward", "tumor.ref_reads.reverse",
                  "tumor.var_reads.forward", "tumor.var_reads.reverse"),	
           sep = ",",	convert = TRUE,
           remove = FALSE)	
# apply the strand filter
calls <- filter(calls, 
                        !(as.numeric(tumor.var_reads.forward) > as.numeric(0.9*as.numeric(tumor.var_reads))) & 
                          !(as.numeric(tumor.var_reads.reverse) > as.numeric(0.9*as.numeric(tumor.var_reads))))
# update actual tumor depth
calls$tumor.depth <- calls$tumor.ref_reads.forward + calls$tumor.ref_reads.reverse +
  calls$tumor.var_reads.forward + calls$tumor.var_reads.reverse

# remove columns for strand bias check since they are no longer needed
calls$tumor.ref_reads.forward <- NULL
calls$tumor.ref_reads.reverse <- NULL
calls$tumor.var_reads.forward <- NULL
calls$tumor.var_reads.reverse <- NULL
calls$tumor.depth4 <- NULL

# Filters that are applied to all samples
# ---------------------------------------

calls <- filter(calls, as.numeric(calls$tumor.allele_freq) >= MIN_VAF)
calls <- filter(calls, as.numeric(calls$tumor.var_reads) >= MIN_VAR_READS_TUMOR)

# keep only variants that have a protein-altering impact
effects.exclude <- c("intron_variant", "intragenic_variant", "", "sequence_feature", "synonymous_variant")
calls <- filter(calls, !(calls$effect %in% effects.exclude))
calls <- filter(calls, calls$impact != "LOW") # 1336 of 1797 
calls <- filter(calls, calls$biotype %in% biotype.keeps) # same
calls <- unique(calls[grep("^ENST", calls$feature_id), ]) # 1053 of 1458

# remove duplicates (e.g. due to multiple effects)
calls <- unique(calls)	

calls$id = as.character(paste(	
  calls$chr, 	
  calls$pos, 	
  calls$ref,	
  calls$alt,	
  sep = "."	
  ))

write.table(calls, "~/Documents/CLL_WGS/varScan/varScan-snvs_indels-min_VAF_5.txt", 
            sep = "\t", quote = FALSE, row.names = FALSE)
```

#table(calls$sample)
MB0141CLL-tumor1_MB0141CLL-buccal MB0141CLL-tumor2_MB0141CLL-buccal 
                              409                               382 
MB0291CLL-tumor1_MB0291CLL-buccal MB0291CLL-tumor2_MB0291CLL-buccal 
                              228                               220 
MB0141CLL-buccal_MB0141CLL-normal MB0141CLL-tumor1_MB0141CLL-normal 
                              891                               206 
MB0141CLL-tumor2_MB0141CLL-normal MB0291CLL-buccal_MB0291CLL-normal 
                              209                               459 
MB0291CLL-tumor1_MB0291CLL-normal MB0291CLL-tumor2_MB0291CLL-normal 
                              189                               226 
                              
## Visualization of numbers of somatic mutations per pair

```{r}
calls <- read.table("~/Documents/CLL_WGS/calls/no_normal_or_UTR_filter/varScan-snvs_indels-min_VAF_5.txt", sep = "\t", header = TRUE, fill = TRUE)

##########################################################
## NEW: apply filters for (1) normal contamination and (2) to remove UTR mutations
##########################################################

utr.effects <- c("3_prime_UTR_variant", "5_prime_UTR_variant")
calls.filtered <- filter(calls, as.numeric(calls$normal.var_reads) < 2 & !(calls$effect %in% utr.effects))

calls <- calls.filtered

##########################################################

plot_data <- as.data.frame(table(calls$sample, calls$type))
colnames(plot_data) <- c("sample", "type", "num_mutations")

p <- ggplot(data = plot_data, aes(x = reorder(sample, -num_mutations), y = num_mutations, fill = type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("") + ylab("Number of Mutations")

p.flip <- ggplot(data = plot_data, aes(x = reorder(sample, -num_mutations), y = num_mutations, fill = type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  coord_flip()

pdf("~/Documents/CLL_WGS/figures/barplot-num_mutations-type.pdf", width = 7, height = 5)
p.flip
dev.off()

# make a pie chart of indels in sample with high indel count
temp <- filter(calls, calls$sample == "MB0141CLL-buccal_MB0141CLL-normal")
temp.indels <- filter(temp, temp$type == "indel")
temp.indels$effect <- factor(temp.indels$effect)
indel.plot_data <- as.data.frame(table(temp.indels$effect))
colnames(indel.plot_data) <- c("effect", "num_mutations")

bp <- ggplot(indel.plot_data, aes(x="", y=num_mutations, fill=effect)) +
  geom_bar(width = 1, stat = "identity") + xlab("")

pie <- bp + coord_polar("y", start = 0) + 
  scale_fill_brewer(palette="Pastel1") + 
  theme(legend.position="bottom")

pdf("~/Documents/CLL_WGS/figures/pie-MB0141CLL-buccal_MB0141CLL-normal-indels-effect.pdf", width = 7, height = 6)
pie
dev.off()

```

## Look at allele frequency distributions for MB0141

```{r}
# MB0141 calls
calls.MB0141.filtered <- calls[! grepl("MB0291", calls$sample), ]
# remove buccal vs. normal
calls.MB0141.filtered <- filter(calls.MB0141.filtered, calls.MB0141.filtered$sample != "MB0141CLL-buccal_MB0141CLL-normal")
calls.MB0141.filtered$sample <- factor(calls.MB0141.filtered$sample)

samples <- unique(calls.MB0141.filtered$sample)

# for each sample, create AF density plot with vertical at peak value
AF.plots <- vector('list', length(samples))

# Histogram with density plot
for (i in seq_along(samples)) {
  temp <- subset(calls.MB0141.filtered, calls.MB0141.filtered$sample == samples[i])
  
  num_mutations <- nrow(temp)
  
  # identify peak value for current sample
  max_y_density <- which.max(density(temp$tumor.allele_freq)$y)
  max_x_intercept <- density(temp$tumor.allele_freq)$x[max_y_density]
  max_y_intercept <- density(temp$tumor.allele_freq)$y[max_y_density]
  
  print(max_y_intercept)
  
  AF.plots[[i]] <- ggplot(temp, aes(as.numeric(tumor.allele_freq))) + 
    geom_histogram(aes(y=..density..), colour="black", fill="white") +
    geom_density(alpha=.2, fill="#FF6666") +
    geom_vline(xintercept = max_x_intercept, col="red", lty=2) +
    xlab("Allele frequency (%)") + ylab("Density") +
    annotate("text", 
             y = max_y_intercept+0.2*max_y_intercept,
             x = max(temp$tumor.allele_freq),
             label = paste("Peak AF: ", round(max_x_intercept, digits = 2), "%", sep = ""),
             hjust = 1) + 
    theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
    ylim(0, max_y_intercept+0.25*max_y_intercept) +
    ggtitle(label = paste(temp$sample, " (n = ", num_mutations, ")", sep = ""))
}
#do.call("grid.arrange", c(AF.plots, ncol=1))
pdf("~/Documents/CLL_WGS/figures/AF_density-MB0141_tumor_pairs.pdf", width = 10, height = 5)
grid.arrange(grobs = AF.plots, ncol=2, bottom = "Allele frequency (%)", left = "Density")
dev.off()
```

## Look at allele frequency distributions for MB0291

```{r}
# MB0291 calls
calls.MB0291.filtered <- calls[! grepl("MB0141", calls$sample), ]
# remove buccal vs. normal
calls.MB0291.filtered <- filter(calls.MB0291.filtered, calls.MB0291.filtered$sample != "MB0291CLL-buccal_MB0291CLL-normal")
calls.MB0291.filtered$sample <- factor(calls.MB0291.filtered$sample)

samples <- unique(calls.MB0291.filtered$sample)

# for each sample, create AF density plot with vertical at peak value
AF.plots <- vector('list', length(samples))

# Histogram with density plot
for (i in seq_along(samples)) {
  temp <- subset(calls.MB0291.filtered, calls.MB0291.filtered$sample == samples[i])
  
  num_mutations <- nrow(temp)
  
  # identify peak value for current sample
  max_y_density <- which.max(density(temp$tumor.allele_freq)$y)
  max_x_intercept <- density(temp$tumor.allele_freq)$x[max_y_density]
  max_y_intercept <- density(temp$tumor.allele_freq)$y[max_y_density]
  
  print(max_y_intercept)
  
  AF.plots[[i]] <- ggplot(temp, aes(as.numeric(tumor.allele_freq))) + 
    geom_histogram(aes(y=..density..), colour="black", fill="white") +
    geom_density(alpha=.2, fill="#FF6666") +
    geom_vline(xintercept = max_x_intercept, col="red", lty=2) +
    xlab("Allele frequency (%)") + ylab("Density") +
    annotate("text", 
             y = max_y_intercept+0.2*max_y_intercept,
             x = max(temp$tumor.allele_freq),
             label = paste("Peak AF: ", round(max_x_intercept, digits = 2), "%", sep = ""),
             hjust = 1) + 
    theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
    ylim(0, max_y_intercept+0.25*max_y_intercept) +
    ggtitle(label = paste(temp$sample, " (n = ", num_mutations, ")", sep = ""))
}
#do.call("grid.arrange", c(AF.plots, ncol=1))
pdf("~/Documents/CLL_WGS/figures/AF_density-MB0291_tumor_pairs.pdf", width = 10, height = 5)
grid.arrange(grobs = AF.plots, ncol=2, bottom = "Allele frequency (%)", left = "Density")
dev.off()
```

## Examine overlap

Venn diagram 1 - find overlap between:
  a. MB0141CLL-tumor1_MB0141CLL-buccal
  b. MB0141CLL-tumor1_MB0141CLL-normal
  
Venn diagram 2 - find overlap between:
  a. MB0141CLL-tumor2_MB0141CLL-buccal
  b. MB0141CLL-tumor2_MB0141CLL-normal
  
Venn diagram 3 - find overlap between:
  a. MB0291CLL-tumor1_MB0291CLL-buccal
  b. MB0291CLL-tumor1_MB0291CLL-normal
  
Venn diagram 4 - find overlap between:
  a. MB0291CLL-tumor2_MB0291CLL-buccal
  b. MB0291CLL-tumor2_MB0291CLL-normal


```{r}
library(VennDiagram)

# overlap of variants for each tumor duplicate (variants identified with buccal vs. sorted normal)

MB0141_T1_vs_buccal <- unique(filter(calls, calls$sample == "MB0141CLL-tumor1_MB0141CLL-buccal"))$id
MB0141_T1_vs_normal <- unique(filter(calls, calls$sample == "MB0141CLL-tumor1_MB0141CLL-normal"))$id

venn.diagram(
  x = list(MB0141_T1_vs_buccal, MB0141_T1_vs_normal),
  category.names = c("MB0141_T1_vs_buccal" , "MB0141_T1_vs_normal"),
  filename = "~/Documents/CLL_WGS/figures/venn-MB0141_T1.png",
  output=TRUE
)

MB0141_T2_vs_buccal <- unique(filter(calls, calls$sample == "MB0141CLL-tumor2_MB0141CLL-buccal"))$id
MB0141_T2_vs_normal <- unique(filter(calls, calls$sample == "MB0141CLL-tumor2_MB0141CLL-normal"))$id

venn.diagram(
  x = list(MB0141_T2_vs_buccal, MB0141_T2_vs_normal),
  category.names = c("MB0141_T2_vs_buccal" , "MB0141_T2_vs_normal"),
  filename = "~/Documents/CLL_WGS/figures/venn-MB0141_T2.png",
  output=TRUE
)

MB0291_T1_vs_buccal <- unique(filter(calls, calls$sample == "MB0291CLL-tumor1_MB0291CLL-buccal"))$id
MB0291_T1_vs_normal <- unique(filter(calls, calls$sample == "MB0291CLL-tumor1_MB0291CLL-normal"))$id

venn.diagram(
  x = list(MB0291_T1_vs_buccal, MB0291_T1_vs_normal),
  category.names = c("MB0291_T1_vs_buccal" , "MB0291_T1_vs_normal"),
  filename = "~/Documents/CLL_WGS/figures/venn-MB0291_T1.png",
  output=TRUE
)

MB0291_T2_vs_buccal <- unique(filter(calls, calls$sample == "MB0291CLL-tumor2_MB0291CLL-buccal"))$id
MB0291_T2_vs_normal <- unique(filter(calls, calls$sample == "MB0291CLL-tumor2_MB0291CLL-normal"))$id

venn.diagram(
  x = list(MB0291_T2_vs_buccal, MB0291_T2_vs_normal),
  category.names = c("MB0291_T2_vs_buccal" , "MB0291_T2_vs_normal"),
  filename = "~/Documents/CLL_WGS/figures/venn-MB0291_T2.png",
  output=TRUE
)

# overlap of all variants for each patient

venn.diagram(
  x = list(MB0141_T1_vs_buccal, MB0141_T1_vs_normal, MB0141_T2_vs_buccal, MB0141_T2_vs_normal),
  category.names = c("MB0141_T1_vs_buccal", "MB0141_T1_vs_normal", "MB0141_T2_vs_buccal", "MB0141_T2_vs_normal"),
  filename = "~/Documents/CLL_WGS/figures/venn-MB0141T-all_pairs.png",
  output=TRUE
)

venn.diagram(
  x = list(MB0291_T1_vs_buccal, MB0291_T1_vs_normal, MB0291_T2_vs_buccal, MB0291_T2_vs_normal),
  category.names = c("MB0291_T1_vs_buccal", "MB0291_T1_vs_normal", "MB0291_T2_vs_buccal", "MB0291_T2_vs_normal"),
  filename = "~/Documents/CLL_WGS/figures/venn-MB0291_T2-all_pairs.png",
  output=TRUE
)

```

## Look at specific overlaps within venn for MB0141

```{r}
# print out overlaps
library(gplots)
MB0141.venn <- venn(list(MB0141_T1_vs_buccal, MB0141_T2_vs_buccal, MB0141_T2_vs_normal, MB0141_T1_vs_normal))
#MB0141.table <- print(MB0141.venn)

# get intersections
MB0141.intersect <- attr(MB0141.venn, "intersection")

# Look at all of the subsets
str(MB0141.intersect)

# A = T1 vs. buccal
# B = T2 vs. buccal
# C = T2 vs. sorted normal
# D = T1 vs. sorted normal

# interested in: A unique, B unique, C unique, D unique, A+B, C+D, A+B+C+D

# Extract and combine the subsets of interest..

# variants unique to T1 vs. buccal
A_only <- unique(c(MB0141.intersect$A))
A_only.label <- rep("T1_v_buccal_only", length(A_only))
A_only.df <- data.frame(A_only, A_only.label)
colnames(A_only.df) <- c("id", "set")

# variants unique to T2 vs. buccal
B_only <- unique(c(MB0141.intersect$B))
B_only.label <- rep("T2_v_buccal_only", length(B_only))
B_only.df <- data.frame(B_only, B_only.label)
colnames(B_only.df) <- c("id", "set")

# variants unique to buccal normal
A_and_B <- unique(c(MB0141.intersect$'A:B'))
A_and_B.label <- rep("T1_and_T2_v_buccal", length(A_and_B))
A_and_B.df <- data.frame(A_and_B, A_and_B.label)
colnames(A_and_B.df) <- c("id", "set")

# variants unique to T1 vs. sortedN
C_only <- unique(c(MB0141.intersect$C))
C_only.label <- rep("T2_v_sortedN_only")
C_only.df <- data.frame(C_only, C_only.label)
colnames(C_only.df) <- c("id", "set")

# variants unique to T2 vs. sortedN
D_only <- unique(c(MB0141.intersect$D))
D_only.label <- rep("T1_v_sortedN_only")
D_only.df <- data.frame(D_only, D_only.label)
colnames(D_only.df) <- c("id", "set")

# variants unique to sortedN normal
C_and_D <- unique(c(MB0141.intersect$'C:D'))
C_and_D.label <- rep("T1_and_T2_v_sortedN")
C_and_D.df <- data.frame(C_and_D, C_and_D.label)
colnames(C_and_D.df) <- c("id", "set")

# variants found in all 4 pairs
A_B_C_D <- unique(c(MB0141.intersect$'A:B:C:D'))
A_B_C_D.label <- rep("all_pairs", length(A_B_C_D))
A_B_C_D.df <- data.frame(A_B_C_D, A_B_C_D.label)
colnames(A_B_C_D.df) <- c("id", "set")

all.sets <- rbind(A_only.df, B_only.df, A_and_B.df, C_only.df, D_only.df, C_and_D.df, A_B_C_D.df)

# attach set labels to calls
calls.with_sets <- merge(calls, all.sets, by = "id", all.x = TRUE)

write.table(calls.with_sets, 
            "~/Documents/CLL_WGS/calls/up_to_1_normal_and_UTR_filter/varScan-snvs_indels-MB0141_venn_sets.txt",
           sep = "\t", quote = FALSE, row.names = FALSE)

```

## Analyze recurrently mutated genes in MB0141

```{r}
library(ggplot2)
library(ggpubr)

# remove MB0291 calls
calls.MB0141 <- calls.with_sets[! grepl("MB0291", calls.with_sets$sample), ]

# intersect of all four pairs
calls.intersect <- filter(calls.MB0141, calls.MB0141$set == "all_pairs")
calls.intersect$gene <- factor(calls.intersect$gene)

# intersect of T vs. sorted N
calls.sortedN <- filter(calls.MB0141, calls.MB0141$set == "T1_and_T2_v_sortedN")
calls.sortedN$gene <- factor(calls.sortedN$gene)

# intersect of T vs. buccal
calls.buccal <- filter(calls.MB0141, calls.MB0141$set == "T1_and_T2_v_buccal")
calls.buccal$gene <- factor(calls.buccal$gene)

plot_data1 <- as.data.frame(table(calls.intersect$gene, calls.intersect$type))
colnames(plot_data1) <- c("gene", "type", "num_mutations")
plot_data2 <- as.data.frame(table(calls.sortedN$gene, calls.sortedN$type))
colnames(plot_data2) <- c("gene", "type", "num_mutations")
plot_data3 <- as.data.frame(table(calls.buccal$gene, calls.buccal$type))
colnames(plot_data3) <- c("gene", "type", "num_mutations")

p1 <- ggplot(data = plot_data1, aes(x = reorder(gene, num_mutations), y = num_mutations, fill=type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  theme(axis.text.y = element_text(face="italic")) +
  coord_flip() + ggtitle("All pairs (22)") 
p2 <- ggplot(data = plot_data2, aes(x = reorder(gene, num_mutations), y = num_mutations, fill=type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  theme(axis.text.y = element_text(face="italic")) +
  coord_flip() + ggtitle("T vs. sortedN (5)")
p3 <- ggplot(data = plot_data3, aes(x = reorder(gene, num_mutations), y = num_mutations, fill=type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  theme(axis.text.y = element_text(face="italic")) +
  coord_flip() + ggtitle("T vs. buccal (49)")

pdf("~/Documents/CLL_WGS/figures/barplot-MB0141-all_pairs-num_mutations_by_gene.pdf", width = 12, height = 5)
ggarrange(p1, p2, p3, ncol=3, common.legend = TRUE, legend="bottom")
dev.off()

# T1 vs. buccal only
calls.T1_buccal <- filter(calls.MB0141, calls.MB0141$set == "T1_v_buccal_only")
calls.T1_buccal$gene <- factor(calls.T1_buccal$gene)

# T2 vs. buccal only
calls.T2_buccal <- filter(calls.MB0141, calls.MB0141$set == "T2_v_buccal_only")
calls.T2_buccal$gene <- factor(calls.T2_buccal$gene)

# T1 vs. sorted N only
calls.T1_sortedN <- filter(calls.MB0141, calls.MB0141$set == "T1_v_sortedN_only")
calls.T1_sortedN$gene <- factor(calls.T1_sortedN$gene)

# T2 vs. sorted N only
calls.T2_sortedN <- filter(calls.MB0141, calls.MB0141$set == "T2_v_sortedN_only")
calls.T2_sortedN$gene <- factor(calls.T2_sortedN$gene)

plot_data4 <- as.data.frame(table(calls.T1_buccal$gene, calls.T1_buccal$type))
colnames(plot_data4) <- c("gene", "type", "num_mutations")
#plot_data4 <- filter(plot_data4, as.numeric(plot_data4$num_mutations) > 1)
plot_data5 <- as.data.frame(table(calls.T2_buccal$gene, calls.T2_buccal$type))
colnames(plot_data5) <- c("gene", "type", "num_mutations")
#plot_data5 <- filter(plot_data5, as.numeric(plot_data5$num_mutations) > 1)
plot_data6 <- as.data.frame(table(calls.T1_sortedN$gene, calls.T1_sortedN$type))
colnames(plot_data6) <- c("gene", "type", "num_mutations")
#plot_data6 <- filter(plot_data6, as.numeric(plot_data6$num_mutations) > 1)
plot_data7 <- as.data.frame(table(calls.T2_sortedN$gene, calls.T2_sortedN$type))
colnames(plot_data7) <- c("gene", "type", "num_mutations")
#plot_data7 <- filter(plot_data7, as.numeric(plot_data7$num_mutations) > 1)

p4 <- ggplot(data = plot_data4, aes(x = reorder(gene, num_mutations), y = num_mutations, fill=type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  theme(axis.text.y = element_text(face="italic")) +
  coord_flip() + ggtitle("T1 vs. buccal only (64)") 
p5 <- ggplot(data = plot_data5, aes(x = reorder(gene, num_mutations), y = num_mutations, fill=type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  theme(axis.text.y = element_text(face="italic")) +
  coord_flip() + ggtitle("T2 vs. buccal only (66)") 
p6 <- ggplot(data = plot_data6, aes(x = reorder(gene, num_mutations), y = num_mutations, fill=type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  theme(axis.text.y = element_text(face="italic")) +
  coord_flip() + ggtitle("T1 vs. sortedN only (20)") 
p7 <- ggplot(data = plot_data7, aes(x = reorder(gene, num_mutations), y = num_mutations, fill=type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  theme(axis.text.y = element_text(face="italic")) +
  coord_flip() + ggtitle("T2 vs. sortedN only (23)") 

pdf("~/Documents/CLL_WGS/figures/barplot-MB0141-unique_mutations-by_gene.pdf", width = 14, height = 6)
ggarrange(p4, p5, p6, p7, ncol=4, common.legend = TRUE, legend="bottom")
dev.off()

```

## Look at specific overlaps within venn for MB0291

```{r}
# print out overlaps
library(gplots)
MB0291.venn <- venn(list(MB0291_T1_vs_buccal, MB0291_T2_vs_buccal, MB0291_T2_vs_normal, MB0291_T1_vs_normal))
#MB0291.table <- print(MB0291.venn)

# get intersections
MB0291.intersect <- attr(MB0291.venn, "intersection")

# Look at all of the subsets
str(MB0291.intersect)

# A = T1 vs. buccal
# B = T2 vs. buccal
# C = T2 vs. sorted normal
# D = T1 vs. sorted normal

# interested in: A unique, B unique, C unique, D unique, A+B, C+D, A+B+C+D

# Extract and combine the subsets of interest..

# variants unique to T1 vs. buccal
A_only <- unique(c(MB0291.intersect$A))
A_only.label <- rep("T1_v_buccal_only", length(A_only))
A_only.df <- data.frame(A_only, A_only.label)
colnames(A_only.df) <- c("id", "set")

# variants unique to T2 vs. buccal
B_only <- unique(c(MB0291.intersect$B))
B_only.label <- rep("T2_v_buccal_only", length(B_only))
B_only.df <- data.frame(B_only, B_only.label)
colnames(B_only.df) <- c("id", "set")

# variants unique to buccal normal
A_and_B <- unique(c(MB0291.intersect$'A:B'))
A_and_B.label <- rep("T1_and_T2_v_buccal", length(A_and_B))
A_and_B.df <- data.frame(A_and_B, A_and_B.label)
colnames(A_and_B.df) <- c("id", "set")

# variants unique to T1 vs. sortedN
C_only <- unique(c(MB0291.intersect$C))
C_only.label <- rep("T2_v_sortedN_only")
C_only.df <- data.frame(C_only, C_only.label)
colnames(C_only.df) <- c("id", "set")

# variants unique to T2 vs. sortedN
D_only <- unique(c(MB0291.intersect$D))
D_only.label <- rep("T1_v_sortedN_only")
D_only.df <- data.frame(D_only, D_only.label)
colnames(D_only.df) <- c("id", "set")

# variants unique to sortedN normal
C_and_D <- unique(c(MB0291.intersect$'C:D'))
C_and_D.label <- rep("T1_and_T2_v_sortedN")
C_and_D.df <- data.frame(C_and_D, C_and_D.label)
colnames(C_and_D.df) <- c("id", "set")

# variants found in all 4 pairs
A_B_C_D <- unique(c(MB0291.intersect$'A:B:C:D'))
A_B_C_D.label <- rep("all_pairs", length(A_B_C_D))
A_B_C_D.df <- data.frame(A_B_C_D, A_B_C_D.label)
colnames(A_B_C_D.df) <- c("id", "set")

all.sets <- rbind(A_only.df, B_only.df, A_and_B.df, C_only.df, D_only.df, C_and_D.df, A_B_C_D.df)

# attach set labels to calls
calls.with_sets <- merge(calls, all.sets, by = "id", all.x = TRUE)
```

## Analyze recurrently mutated genes in MB0291

```{r}
library(ggplot2)
library(ggpubr)

# remove MB0141 calls
calls.MB0291 <- calls.with_sets[! grepl("MB0141", calls.with_sets$sample), ]

# intersect of all four pairs
calls.intersect <- filter(calls.MB0291, calls.MB0291$set == "all_pairs")
calls.intersect$gene <- factor(calls.intersect$gene)

# intersect of T vs. sorted N
calls.sortedN <- filter(calls.MB0291, calls.MB0291$set == "T1_and_T2_v_sortedN")
calls.sortedN$gene <- factor(calls.sortedN$gene)

# intersect of T vs. buccal
calls.buccal <- filter(calls.MB0291, calls.MB0291$set == "T1_and_T2_v_buccal")
calls.buccal$gene <- factor(calls.buccal$gene)

plot_data1 <- as.data.frame(table(calls.intersect$gene, calls.intersect$type))
colnames(plot_data1) <- c("gene", "type", "num_mutations")
plot_data2 <- as.data.frame(table(calls.sortedN$gene, calls.sortedN$type))
colnames(plot_data2) <- c("gene", "type", "num_mutations")
plot_data3 <- as.data.frame(table(calls.buccal$gene, calls.buccal$type))
colnames(plot_data3) <- c("gene", "type", "num_mutations")

p1 <- ggplot(data = plot_data1, aes(x = reorder(gene, num_mutations), y = num_mutations, fill=type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  theme(axis.text.y = element_text(face="italic")) +
  coord_flip() + ggtitle("All pairs (9)") 
p2 <- ggplot(data = plot_data2, aes(x = reorder(gene, num_mutations), y = num_mutations, fill=type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  theme(axis.text.y = element_text(face="italic")) +
  coord_flip() + ggtitle("T vs. sortedN (17)")
p3 <- ggplot(data = plot_data3, aes(x = reorder(gene, num_mutations), y = num_mutations, fill=type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  theme(axis.text.y = element_text(face="italic")) +
  coord_flip() + ggtitle("T vs. buccal (54)")

pdf("~/Documents/CLL_WGS/figures/barplot-MB0291-all_pairs-num_mutations_by_gene.pdf", width = 12, height = 5)
ggarrange(p1, p2, p3, ncol=3, common.legend = TRUE, legend="bottom")
dev.off()

# T1 vs. buccal only
calls.T1_buccal <- filter(calls.MB0291, calls.MB0291$set == "T1_v_buccal_only")
calls.T1_buccal$gene <- factor(calls.T1_buccal$gene)

# T2 vs. buccal only
calls.T2_buccal <- filter(calls.MB0291, calls.MB0291$set == "T2_v_buccal_only")
calls.T2_buccal$gene <- factor(calls.T2_buccal$gene)

# T1 vs. sorted N only
calls.T1_sortedN <- filter(calls.MB0291, calls.MB0291$set == "T1_v_sortedN_only")
calls.T1_sortedN$gene <- factor(calls.T1_sortedN$gene)

# T2 vs. sorted N only
calls.T2_sortedN <- filter(calls.MB0291, calls.MB0291$set == "T2_v_sortedN_only")
calls.T2_sortedN$gene <- factor(calls.T2_sortedN$gene)

plot_data4 <- as.data.frame(table(calls.T1_buccal$gene, calls.T1_buccal$type))
colnames(plot_data4) <- c("gene", "type", "num_mutations")
# plot_data4 <- filter(plot_data4, as.numeric(plot_data4$num_mutations) > 1)
plot_data5 <- as.data.frame(table(calls.T2_buccal$gene, calls.T2_buccal$type))
colnames(plot_data5) <- c("gene", "type", "num_mutations")
# plot_data5 <- filter(plot_data5, as.numeric(plot_data5$num_mutations) > 1)
plot_data6 <- as.data.frame(table(calls.T1_sortedN$gene, calls.T1_sortedN$type))
colnames(plot_data6) <- c("gene", "type", "num_mutations")
# plot_data6 <- filter(plot_data6, as.numeric(plot_data6$num_mutations) > 1)
plot_data7 <- as.data.frame(table(calls.T2_sortedN$gene, calls.T2_sortedN$type))
colnames(plot_data7) <- c("gene", "type", "num_mutations")
# plot_data7 <- filter(plot_data7, as.numeric(plot_data7$num_mutations) > 1)

p4 <- ggplot(data = plot_data4, aes(x = reorder(gene, num_mutations), y = num_mutations, fill=type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  theme(axis.text.y = element_text(face="italic")) +
  coord_flip() + ggtitle("T1 vs. buccal only (42)") 
p5 <- ggplot(data = plot_data5, aes(x = reorder(gene, num_mutations), y = num_mutations, fill=type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  theme(axis.text.y = element_text(face="italic")) +
  coord_flip() + ggtitle("T2 vs. buccal only (22)") 
p6 <- ggplot(data = plot_data6, aes(x = reorder(gene, num_mutations), y = num_mutations, fill=type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  theme(axis.text.y = element_text(face="italic")) +
  coord_flip() + ggtitle("T1 vs. sortedN only (25)") 
p7 <- ggplot(data = plot_data7, aes(x = reorder(gene, num_mutations), y = num_mutations, fill=type)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  xlab("") + ylab("Number of Mutations") +
  theme(axis.text.y = element_text(face="italic")) +
  coord_flip() + ggtitle("T2 vs. sortedN only (20)") 

pdf("~/Documents/CLL_WGS/figures/barplot-MB0291-unique_mutations-by_gene.pdf", width = 14, height = 6)
ggarrange(p4, p5, p6, p7, ncol=4, common.legend = TRUE, legend="bottom")
dev.off()

```


## Print out full set of calls combining set labels for both MB0141 + MB0291

```{r}
calls.all <- rbind(calls.MB0141, calls.MB0291)

write.table(calls.all, 
            "~/Documents/CLL_WGS/calls/up_to_1_normal_and_UTR_filter/varScan-snvs_indels-min_VAF_5-both_set_labels.txt",
           sep = "\t", quote = FALSE, row.names = FALSE)

```

## Apply two new filters for (1) normal contamination and (2) to remove UTR mutations

```{r}


```


